Sparse Function-space Representation \\ of Neural Networks

### 2023-05-24 paper/main.tex
fixed small typos

\subsection{Computational complexity}
\label{app:comp_compl}
% Big O
% train neural network
% compute the NTK -> from the NTK paper
% we outer product of the kzi O(NM^2)
% inversion O(M^3)
In this section, we analyze the computational complexity of \our. We first need to train a neural network $f_\weights$ and then compute the posterior covariance matrix that characterizes our sparse GP representation of the model. The first step then requires the usual complexity cost for training a neural network, where defining the cost of a forward pass (and backward pass
%\todo{these two should be similar, we can't cite the NTK paper were they are using this notation unfortunately}
) as $[F\!P]$, this can roughly sketched as $\mathcal{O}(2  E  N[F\!P])$, where $E$ is the number of training epochs and  $N$ the number of training samples. Thus, there is no additional cost involved in the training phase of the neural network.

Given the trained network, we construct the covariance function  which is defined through the NTK~\citep{jacot2018neural}. We relied on the Jacobian contraction implementation using PyTorch~\citep{paszke2019pytorch}. To compute the kernel Gram matrix for a batch of $B$ samples with this implementation has a computational complexity of $\mathcal{O}( B C [F\!P] + B^2 C^2 P)$, where $C$ is the number of NN outputs and $P$ is the number of parameters of the neural network. In \cref{eq:dual_sparse}, we compute $\MB_\vu$ iterating through all the $N$ training points to compute the $\vkzi$ and then use them to compute the outer product, that has a cost of $\mathcal{O}(M^2)$, with $M$ being the number of inducing points following the notation used in the main paper. The total cost of this operation becomes $\mathcal{O}(N M^2)$, which is also the dominating cost of our approach.

% Practical implementation
% mention that we do that in batches not full N 
% role of jitter 
% role of clamping
% GPU A100 Nvidia
In practical terms, some care needs to be taken during the implementation, both in terms of memory consumption and numerical stability. For the former, we batched over the number of training samples, $N$, which keeps the memory consumption in check without need for any further approximations. The computation of the terms in \cref{eq:dual_sparse_post} require the inversion of an $M \times M$ matrix, that we implemented in a numerically stable manner by using Cholesky factorization. We also introduce a small jitter and clip of the GGN matrix for numerical stability where needed (see discussions in \cref{app:experiments}). 
 We run our experiments on GPU clusters equipped with NVIDIA A100 GPUs.

\section{\our for continual learning}
\label{app:cl}
This section provides additional details about the derivation of the \our-based regularizer described in \cref{sec:sequential} and how we extended it for dealing with the multi-output setting in our experiments.
\subsection{Derivation of the regularizer term}
For continual learning, the model cannot access the training data from previous tasks up to task $t-1$, denoted here as $\dataset_\textrm{old} = \bigcup_{s=1}^{t-1} \dataset_s$, and for each new task $t$ receives a new chunk of data $\dataset_\textrm{new} = \dataset_t$. On the other hand, in the supervised learning setting (described in \cref{sec:methods}), the neural network model has access to all the training data at once, \ie, the objective we are trying to optimize is
\begin{equation}
	\weights^* = \argmin_\weights \mathcal{L}(\dataset_\textrm{old} \cup \dataset_\textrm{new}, \weights) = \argmin_\weights \sum_{i \in \dataset_\textrm{old} \cup \dataset_\textrm{new}}\ell(f_\weights(\mathbf{x}_{i}), y_i) + \delta \mathcal{R}(\weights).
\end{equation} 
Our goal in the continual setting is to replace the missing data coming from $\dataset_\textrm{old}$ by resorting to a functional regularizer that accounts for that by forcing the predictions of the neural networks over the inducing points $\mathcal{Z}_t$ to come from $f^{\textrm{old}}_\weights(\vz_i) \sim q^\textrm{old}(\vu),  \forall \vz_i \in \mathcal{Z}_t$. As shown in \citep{khan2017conjugate, adam2021dual}, we can view the dual parameters derived in \cref{sec:sparse-dual-gp} as linked to approximate likelihood terms. Using this insight, we can rewrite the sparse GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:
%
\begin{align} \label{eq:normal}
	q(\vu)
	& \propto  \Norm(\mathbf{0}, \MKzz) \prod_{i \in \dataset_\textrm{old}} \exp \! \left(-\frac{\hat{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 \right) \\
        & \propto p(\vu) \Norm(\tilde{\vy}; \vu, \MB_\vu),
\end{align}
% The first normals are the approximate likelihoods 
where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}_i$.
The sparse GP posterior $q(\vu)$ is proportional to the product between the prior over the inducing points $p(\vu)$ and a Gaussian term $\Norm(\tilde{\vy}; \vu, \MB_\vu)$, which arises from the sparse approximation to the likelihood. Replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$, we can see the similarity between the sparse GP and a linear model, for which the kernel is the linear kernel. Therefore, we use the quadratic term from  $\Norm(\tilde{\vy}; \vu, \MB_\vu)$ as our regularizer in place of the missing data term. By taking the $\log(\cdot)$, the regularizer becomes $(\tilde{\vy} - \vu)^\top \bar{\MB}^{-1}(\tilde{\vy} - \vu)$.  Given that $\tilde{\vy}$ are in the function space we can instead us simply the previous neural network predictions $f^{\textrm{old}}_\weights(\vz_i)$ and save computations.

@@

\subsection{Supervised learning on image data sets}
\label{app:image}
% Each experiment with the image data sets consisted of the following steps.
% First, we trained a neural network (MLP for MNIST, CNN for FMNIST) using a prior precision $\delta$.
% After training, we performed inference by calculating the dual parameters.
% Finally, we evaluated the predictive performance of \our on the validation set.
% Similarly to the supervised UCI benchmark tasks, a single image experiment consists of the following steps: first, we train a neural network (a MLP for MNIST, a CNN for FMNIST) in the classification task, using a prior precision $\delta$. After training, we prepare the \our method for predictions by computing the dual parameters, and then evaluate the validation set predictive performance of \our. While in principle our method's predictions consist of the GP mean and variance, we include results for using the neural network prediction as a mean instead. For comparison, we gather the GP subset method results for both the GP and neural network means. This experiment set-up is then replicated for a range of prior precision values and learning rates, as we perform a grid search over these hyperparameters by optimizing for the validation set NLPD. We run the experiment on $5$ seeds and pick the hyperparameter by the best mean performance over the seeds. The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$, and a prior precision of $\delta = 0.0013$ for FMNIST and $\delta = 0.0001$ for MNIST \todo{double check}.

In \cref{sec:image} in the main paper, we included experiments on image data sets. For the image-based classification experiments, we include results using both the GP mean prediction and the NN prediction for both \our and the GP subset.
For the neural network training, we used the same set-up as in \cite{immer2021improving},
training the neural network using Adam for $500$ epochs with a batch size of $512$.
The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$
and we ran each experiment for $5$ seeds.
The MNIST experiments used a prior precision of $\delta=0.0008$
and the FMNIST experiments used a prior precision of $\delta=0.0013$.
The experiments with MNIST  used an MLP with two hidden layers of width 128 and Tanh activations.
In the FMNIST experiments we followed \cite{immer2021improving} and used a CNN
architecture with two fully connected classification layers and Tanh activations.

>>>>

\subsection{Computational complexity}
\label{app:comp_compl}
% Big O
% train neural network
% compute the NTK -> from the NTK paper
% we outer product of the kzi O(NM^2)
% inversion O(M^3)
In this section, we analyze the computational complexity of \our. We first need to train a neural network $f_\weights$ and then compute the posterior covariance matrix that characterizes our sparse GP representation of the model. The first step then requires the usual complexity cost for training a neural network, where defining the cost of a forward pass (and backward pass
%\todo{these two should be similar, we can't cite the NTK paper were they are using this notation unfortunately}
) as $[F\!P]$, this can be roughly sketched as $\mathcal{O}(2  E  N[F\!P])$, where $E$ is the number of training epochs and  $N$ the number of training samples. Thus, there is no additional cost involved in the training phase of the neural network.

Given the trained network, we construct the covariance function, which is defined through the NTK~\citep{jacot2018neural}. We relied on the Jacobian contraction formulation implemented using PyTorch~\citep{paszke2019pytorch}. Computing the kernel Gram matrix for a batch of $B$ samples with this implementation has a computational complexity of $\mathcal{O}( B C [F\!P] + B^2 C^2 P)$, where $C$ is the number of outputs and $P$ is the number of parameters of the neural network. In \cref{eq:dual_sparse}, we compute $\MB_\vu$ iterating through all the $N$ training points to compute all the $\vkzi$. Then, we use them to compute the outer product, which has a cost of $\mathcal{O}(M^2)$, with $M$ being the number of inducing points. The total cost of this operation becomes $\mathcal{O}(N M^2)$, which is also the dominating cost of our approach.

% Practical implementation
% mention that we do that in batches not full N 
% role of jitter 
% role of clamping
% GPU A100 Nvidia
In practical terms, some care needs to be taken during the implementation, both regarding memory consumption and numerical stability. For the former, we batched over the number of training samples, $N$, which keeps the memory consumption in check without the need for any further approximations. The computation of the terms in \cref{eq:dual_sparse_post} requires the inversion of an $M \times M$ matrix, that we implemented in a numerically stable manner using Cholesky factorization. We also introduce a small jitter and clip the GGN matrix for numerical stability where needed (see discussions in \cref{app:experiments}). 
 We run our experiments on a cluster equipped with NVIDIA A100 GPUs.

\section{\our for continual learning}
\label{app:cl}
This section provides additional details about the derivation of the \our-based regularizer described in \cref{sec:sequential} and how we extended it for dealing with the multi-output setting in our experiments.
\subsection{Derivation of the regularizer term}
For continual learning, the model cannot access the training data from previous tasks up to task $t-1$, denoted here as $\dataset_\textrm{old} = \bigcup_{s=1}^{t-1} \dataset_s$, and for each new task $t$ receives a new chunk of data $\dataset_\textrm{new} = \dataset_t$. In contrast, in the supervised learning setting (described in \cref{sec:methods}), the neural network model has access to all the training data at once, and then the objective we are trying to optimize is
\begin{equation}
	\weights^* = \argmin_\weights \mathcal{L}(\dataset_\textrm{old} \cup \dataset_\textrm{new}, \weights) = \argmin_\weights \sum_{i \in \dataset_\textrm{old} \cup \dataset_\textrm{new}}\ell(f_\weights(\mathbf{x}_{i}), y_i) + \delta \mathcal{R}(\weights).
\end{equation} 
Our goal in the continual setting is to replace the missing data coming from $\dataset_\textrm{old}$ by resorting to a functional regularizer that accounts for that by forcing the predictions of the neural networks over the inducing points $\mathcal{Z}_t$ to come from $f^{\textrm{old}}_\weights(\vz_i) \sim q^\textrm{old}(\vu),  \forall \vz_i \in \mathcal{Z}_t$. As shown in \citep{khan2017conjugate, adam2021dual}, we can view the dual parameters derived in \cref{sec:sparse-dual-gp} as linked to approximate likelihood terms. Using this insight, we can rewrite the sparse GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:
%
\begin{align} \label{eq:normal}
	q(\vu)
	& \propto  \Norm(\mathbf{0}, \MKzz) \prod_{i \in \dataset_\textrm{old}} \exp \! \left(-\frac{\hat{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 \right) \\
        & \propto p(\vu) \Norm(\tilde{\vy}; \vu, \MB_\vu),
\end{align}
% The first normals are the approximate likelihoods 
where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}_i$.
The sparse GP posterior $q(\vu)$ is proportional to the product between the prior over the inducing points $p(\vu)$ and a Gaussian term $\Norm(\tilde{\vy}; \vu, \MB_\vu)$, which arises from the sparse approximation to the likelihood. Replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$, we can see the similarity between the sparse GP and a linear model, for which the kernel is the linear kernel. Therefore, we use the quadratic term from  $\Norm(\tilde{\vy}; \vu, \MB_\vu)$ as our regularizer in place of the missing data term. By taking the $\log$, the regularizer becomes $(\tilde{\vy} - \vu)^\top \bar{\MB}^{-1}(\tilde{\vy} - \vu)$.  Given that $\tilde{\vy}$ are in the function space, we can instead use the previous neural network predictions $f^{\textrm{old}}_\weights(\vz_i)$ and save computations.

@@

\subsection{Supervised learning on image data sets}
\label{app:image}
% Each experiment with the image data sets consisted of the following steps.
% First, we trained a neural network (MLP for MNIST, CNN for FMNIST) using a prior precision $\delta$.
% After training, we performed inference by calculating the dual parameters.
% Finally, we evaluated the predictive performance of \our on the validation set.
% Similarly to the supervised UCI benchmark tasks, a single image experiment consists of the following steps: first, we train a neural network (a MLP for MNIST, a CNN for FMNIST) in the classification task, using a prior precision $\delta$. After training, we prepare the \our method for predictions by computing the dual parameters, and then evaluate the validation set predictive performance of \our. While in principle our method's predictions consist of the GP mean and variance, we include results for using the neural network prediction as a mean instead. For comparison, we gather the GP subset method results for both the GP and neural network means. This experiment set-up is then replicated for a range of prior precision values and learning rates, as we perform a grid search over these hyperparameters by optimizing for the validation set NLPD. We run the experiment on $5$ seeds and pick the hyperparameter by the best mean performance over the seeds. The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$, and a prior precision of $\delta = 0.0013$ for FMNIST and $\delta = 0.0001$ for MNIST \todo{double check}.
In \cref{sec:image} in the main paper, we included experiments on image data sets. For the image-based classification experiments, we include results using both the GP mean prediction and the NN prediction for both \our and the GP subset.
For the neural network training, we used the same set-up as in \cite{immer2021improving},
training the neural network using Adam for $500$ epochs with a batch size of $512$.
The results reported in \cref{tbl:imagesuper} used a learning rate of $10^{-4}$, and we ran each experiment for $5$ seeds.
The MNIST experiments used a prior precision of $\delta=0.0008$, and the FMNIST experiments used a prior precision of $\delta=0.0013$.
The experiments with MNIST  used an MLP with two hidden layers of width 128 and Tanh activations.
In the FMNIST experiments, we followed \cite{immer2021improving} and used a CNN architecture composed of three convolutional layers with ReLU activations and three fully connected classification layers with Tanh activations.
